---
title: "Hierarchical model construction"
fontsize: 7pt
output:
  beamer_presentation:
    colortheme: "spruce"
    fonttheme: "structurebold"
    latex_engine: xelatex
header-includes: 
- \usepackage{listings}
- \lstset{basicstyle=\small}
- \setmonofont[Scale=MatchLowercase]{Courier}
- \setmonofont[Scale=0.8]{Courier}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
par(bty='n')
```

# 

1. Classic hierarchical Bayesian models
2. Gaussian process models
3. Projects

# Occupancy models

$$z_i \sim Bernoulli(\psi_i)$$

$$y_{ij} \sim Bernoulli(z_i p)$$

Sites $i, ..., N$

Repeat visits $j, ..., J$

# N-mixture models

$$N_i \sim Poisson(\lambda_i)$$

$$y_{ij} \sim Binomial(N_i, p)$$

# Error in variables models

$$y_i \sim Normal(\alpha + \beta \tilde{x}_i, \sigma_y)$$

$$x_i \sim Normal(\tilde{x}_i, \sigma_x)$$

# Example: modeling poop at ponds

$$y_i \sim Poisson(\mu_i)$$

$$\mu_i = \alpha_0 + log(\tilde{\pi})$$

$$\pi \sim N(\tilde{\pi}, \sigma_\pi)$$

# Zero inflated Poisson

$$p(y_i|\theta,\lambda)
= 
\begin{cases}
\theta + (1 - \theta) Poisson(0 \mid \lambda) & \mbox{if } y = 0 \\
(1 - \theta) Poisson(y_i \mid \lambda) & \mbox{if } y > 0
\end{cases}$$

$\theta$: mixing parameter 

# Zero inflated gamma

$$p(y_i|\theta,\lambda)
= 
\begin{cases}
\theta & \mbox{if } y = 0 \\
(1 - \theta) Gamma(y_i \mid \alpha, \beta) & \mbox{if } y > 0
\end{cases}$$

$\theta$: mixing parameter 

# Beta glm

$$y_i \sim Beta(\alpha, \beta)$$

$$\alpha = \mu \phi$$

$$\beta = (1 - \mu) \phi$$

$$logit(\mu) = X \beta$$

# Hierarchical Bayesian structural equation models

![](fig2.pdf)

# Background: univariate normal

$$x \sim N(\mu, \sigma^2)$$

```{r, echo=FALSE, fig.height=4, fig.width=5}
x <- seq(-3, 3, .1)
plot(x, dnorm(x), type='l', main='Normal(0, 1) probability density')
```

# Background: multivariate normal

$$\boldsymbol{x} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

$\boldsymbol{\mu}$: vector of means

$\boldsymbol{\Sigma}$: covariance matrix

# Bivariate normal probability density

$$\boldsymbol{x} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

```{r, fig.width=8, fig.height=4}
# lets first simulate a bivariate normal sample
library(MASS)
bivn <- mvrnorm(100000, mu = c(0, 0), Sigma = matrix(c(1, .5, .5, 1), 2))

# now we do a kernel density estimate
bivn.kde <- kde2d(bivn[,1], bivn[,2], n = 50)

par(mfrow=c(1, 2))
# now plot your results
persp(bivn.kde, phi = 45, theta = 30, xlab='x1', ylab='x2', zlab='p(x1, x2)')

# fancy contour with image
image(bivn.kde, xlab='x1', ylab='x2'); contour(bivn.kde, add = T)
```

# Bivariate normal parameters

$\boldsymbol{\mu} = \begin{bmatrix}
\mu_1 \\
\mu_2
\end{bmatrix},$
$\boldsymbol{\Sigma} = \begin{bmatrix}
Cov[X_1, X_1] & Cov[X_1, X_2] \\
Cov[X_2, X_1] & Cov[X_2, X_2]
\end{bmatrix}$

```{r, fig.width=6, fig.height=6}
Sigma <- matrix(c(1, .5, .5, 1), nrow=2)
n <- 10000
z <- matrix(rnorm(n), nrow=nrow(Sigma))
y <- t(chol(Sigma)) %*% z
plot(y[1, ], y[2, ], xlab=expression(x[1]), ylab=expression(x[2]))
text(0, 0, labels=expression(bold(mu)), col='red', cex=2)
```

# Bivariate normal parameters

$\boldsymbol{\mu} = \begin{bmatrix}
0 \\
0
\end{bmatrix},$
$\boldsymbol{\Sigma} = \begin{bmatrix}
1 & 0.5 \\
0.5 & 1
\end{bmatrix}$

```{r, fig.width=6, fig.height=6}
plot(y[1, ], y[2, ], xlab=expression(x[1]), ylab=expression(x[2]))
text(0, 0, labels=expression(bold(mu)), col='red', cex=2)
```

# Uncorrelated bivariate normal

$\boldsymbol{\mu} = \begin{bmatrix}
0 \\
0
\end{bmatrix},$
$\boldsymbol{\Sigma} = \begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}$

```{r, fig.width=6, fig.height=6}
Sigma <- matrix(c(1, 0, 0, 1), nrow=2)
y <- t(chol(Sigma)) %*% z
plot(y[1, ], y[2, ], xlab=expression(x[1]), ylab=expression(x[2]))
text(0, 0, labels=expression(bold(mu)), col='red', cex=2)
```

# Common notation

$\boldsymbol{\mu} = \begin{bmatrix}
\mu_1 \\
\mu_2
\end{bmatrix},$
$\boldsymbol{\Sigma} = \begin{bmatrix}
\sigma_1^2 & \rho \sigma_1 \sigma_2 \\
\rho \sigma_1 \sigma_2 & \sigma_2^2
\end{bmatrix}$

# Common notation

$\boldsymbol{\mu} = \begin{bmatrix}
\mu_1 \\
\mu_2
\end{bmatrix},$
$\boldsymbol{\Sigma} = \begin{bmatrix}
\sigma_1^2 & \rho \sigma_1 \sigma_2 \\
\rho \sigma_1 \sigma_2 & \sigma_2^2
\end{bmatrix}$

$Cov[X_1, X_1] = Var[X_1] = \sigma_1^2$

$Cov[X_1, X_2] = \rho \sigma_1 \sigma_2$

$\Sigma$ must be symmetric and positive semi-definite

# Classic linear modeling

$$y = X\beta + \epsilon$$

$$\epsilon \sim N(0, \sigma^2)$$

Functional form determined by $X \beta$

# Linear model functional forms

e.g. $y = \mu(x) + \epsilon$

```{r}
n <- 20
x <- rnorm(n)
beta <- c(1, 1, -.5)
X <- matrix(c(rep(1, n), x, x^3), ncol=3)
y <- c(scale(X %*% beta + rnorm(n)))
plot(x, y, ylim=range(y) * 2)
m1 <- lm(y ~ x)
m2 <- lm(y ~ x + I(x^2))
abline(m1)
newx <- seq(min(x), max(x), .01)
p1 <- predict(m1, data.frame(x=newx), interval='prediction')
p2 <- predict(m2, data.frame(x=newx), interval='prediction')
matlines(newx, p1, lty=c(1, 2, 2), col=1)
matlines(newx, p2, lty=c(1, 2, 2), col=2)
```

# Why not set a prior on $\mu(x)$?

*Gaussian process* as a prior for $\mu(x)$

$$y \sim N(\mu(x), \sigma^2)$$

$$\mu(x) \sim GP(m, k)$$

# GP prior for $\mu(x)$

$y \sim N(\mu(x), \sigma^2)$

$\mu(x) \sim GP(m, k)$

```{r, fig.width=6, fig.height=4}
library(scales)
plot(x, y, ylim=range(y) * 2, pch=19)
D <- as.matrix(dist(x))
C <- function(sigma, d, rho){
  stopifnot(sigma > 0)
  stopifnot(rho > 0)
  sigma ^ 2 * exp(-d^2 / (2 * rho ^ 2))
}
sigma_e <- .00001
n_p <- 1000
x_p <- sort(runif(n_p, min(x), max(x)))
d_mat <- as.matrix(dist(x_p))
Emat <- diag(rep(sigma_e^2, n_p))
# simulate realizations
for (i in 1:100){
  sigma <- runif(1, 0, 5)
  rho <- runif(1, 0, 4)
  Cmat <- C(sigma, d_mat, rho)
  L_c <- t(chol(Cmat + Emat))
  z <- rnorm(n_p)
  y_p <- L_c %*% z
  lines(x_p, y_p + mean(y), col=alpha(1, .1))
}
title('Data and realizations from a GP prior')
```

# Wait, what's Gaussian about that?

If $\mu(x) \sim GP(m, k)$, then 

$\mu(x_1), ..., \mu(x_n) \sim N(m(x_1), ..., m(x_n), K(x_1, ..., x_n)$

$m$ and $k$ are functions!

# Mean function: m

Classic example: $m(x) = X \beta$

e.g., $\mu(x) \sim GP(X \beta, k(x))$

But, the covariance function $k(x)$ is the real star.

# Covariance functions

$k$ specifies covariance between to $x$ values

Squared exponential covariance:

$$k(x, x') = \tau^2 exp\Big(-\dfrac{|x - x'|^2}{\phi}\Big)$$

Lots of [options](http://www.gaussianprocess.org/gpml/chapters/RW4.pdf): smooth, jaggety, periodic

# Example of squared exponential

$$ \boldsymbol{K} = \begin{bmatrix}
\tau^2 exp(-\frac{|x_1 - x_1|^2}{\phi}) & \tau^2 exp(-\frac{|x_1 - x_2|^2}{\phi}) \\
\tau^2 exp(-\frac{|x_2 - x_1|^2}{\phi}) & \tau^2 exp(-\frac{|x_2 - x_2|^2}{\phi})
\end{bmatrix}$$

# Example of squared exponential

$$ \boldsymbol{K} = \begin{bmatrix}
\tau^2 exp(-\frac{0^2}{\phi}) & \tau^2 exp(-\frac{|x_1 - x_2|^2}{\phi}) \\
\tau^2 exp(-\frac{|x_2 - x_1|^2}{\phi}) & \tau^2 exp(-\frac{0^2}{\phi})
\end{bmatrix}$$

# Example of squared exponential

$$ \boldsymbol{K} = \begin{bmatrix}
\tau^2 exp(0) & \tau^2 exp(-\frac{|x_1 - x_2|^2}{\phi}) \\
\tau^2 exp(-\frac{|x_2 - x_1|^2}{\phi}) & \tau^2 exp(0)
\end{bmatrix}$$

# Example of squared exponential

$$ \boldsymbol{K} = \begin{bmatrix}
\tau^2 & \tau^2 exp(-\frac{|x_1 - x_2|^2}{\phi}) \\
\tau^2 exp(-\frac{|x_2 - x_1|^2}{\phi}) & \tau^2
\end{bmatrix}$$

$Cor(\mu(x_1), \mu(x_2)) = exp(-\frac{|x_1 - x_2|^2}{\phi})$.

# Correlation function

$Cor(\mu(x_1), \mu(x_2)) = exp(-\frac{|x_1 - x_2|^2}{\phi})$.

```{r, echo = FALSE}
d <- seq(0, 1, .01)
sq_exp <- function(d, phi) {
  exp(-d^2 / phi)
}
plot(d, sq_exp(d, .1), type='l', 
     xlab='Distance between x1 & x2', 
     ylab='Correlation: mu(x1) & mu(x2)', 
     col = alpha(1, .1))

for (i in 1:100){
  lines(d, sq_exp(d, abs(rnorm(1, .1, .1))), 
        col = alpha(1, .1))
}
```

# Gaussian process realizations

```{r}
# Simulating Gaussian processes
library(akima)
library(scales)

## Univariate inputs ---------------------------
n <- 1000
x <- sort(runif(n, -10, 10))

# squared exponential --------------------------
l <- 1  # range parameter
dmat <- as.matrix(dist(x))
C <- exp(-dmat^2 / (2 * l^2)) + diag(rep(1E-6, n))

# z %*% L produces multivariate normal draws from MVN(0, Sigma),
# where L %*% t(L) = Sigma. i.e., L is a cholesky decomposition of Sigma
# and z ~ normal(0, 1)
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```

# Gaussian process realizations

```{r}
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```


# Gaussian process realizations

```{r}
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```


# Gaussian process realizations

```{r}
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```


# Gaussian process realizations

```{r}
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```

# Gaussian process with nonzero mean function

$y \sim N(\mu, \sigma_y)$

$\mu \sim GP(X \beta, k(x)$

```{r}
# or, built into a linear model
alpha <- -1
beta <- .5
y <- alpha + beta * x + z %*% chol(C)
plot(x, y, type='l')
```

# Ornstein–Uhlenbeck Gaussian process

$k(x_1, x_2) = e^{\frac{d_{x_1, x_2}}{\phi}}$

```{r}
C <- exp(-dmat / l)
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```

# Periodic Gaussian process

$k(x_1, x_2) = exp(\frac{2 \text{sin}^2 (d / 2)}{\phi})$

```{r}
C <- exp(- 2 * sin(dmat / 2) * sin(dmat / 2) / l) + diag(rep(.000001, n))
z <- rnorm(n)
y <- z %*% chol(C)
plot(x, y, type='l')
```

# Combining Gaussian processes

e.g., sums and products of covariance functions

```{r}
C <- exp(- 2 * sin(dmat / 2) * sin(dmat / 2) / l) + diag(rep(.000001, n))
C2 <- exp(-dmat / l)
z <- rnorm(n)
y <- z %*% chol(2 * C  + .1 * C2)
plot(x, y, type='l')
title('Periodic OU Gaussian process')
```

# Combining Gaussian processes

```{r}
C <- exp(- 2 * sin(dmat / 2) * sin(dmat / 2) / l * 30) + diag(rep(.000001, n))
C2 <- exp(- 2 * sin(dmat / 2) * sin(dmat / 2) / l) + diag(rep(.000001, n))
z <- rnorm(n)
y <- z %*% chol(.1 * C  + 2 * C2)
plot(x, y, type='l')
title('Doubly periodic Gaussian process')
```

# Multidimensional inputs

```{r}
## 2d gaussian process -------------------------------
# squared exponential covariance function:
C <- function(sigma, d, rho){
  stopifnot(sigma > 0)
  stopifnot(rho > 0)
  sigma ^ 2 * exp(-d^2 / (2 * rho ^ 2))
}

sigma <- 1
rho <- .7
sigma_e <- .001
n <- 2000
x1 <- runif(n, 0, 10)
x2 <- runif(n, 0, 10)
d_mat <- as.matrix(dist(cbind(x1, x2)))
Cmat <- C(sigma, d_mat, rho)
Emat <- diag(rep(sigma_e^2, n))
# simulate realizations
L_c <- t(chol(Cmat + Emat))
z <- rnorm(n)
y <- L_c %*% z
s <- interp(x1, x2, y, nx=300, ny=300)
image(s, col=rainbow(100))
title('Squared exponential 2d Gaussian process')
```

# Multidimensional inputs

```{r}
## 2d gaussian process -------------------------------
# OU covariance function:
C <- function(sigma, d, rho){
  stopifnot(sigma > 0)
  stopifnot(rho > 0)
  sigma ^ 2 * exp(-d / (rho))
}

sigma <- 1
rho <- .7
sigma_e <- .001
n <- 2000
x1 <- runif(n, 0, 10)
x2 <- runif(n, 0, 10)
d_mat <- as.matrix(dist(cbind(x1, x2)))
Cmat <- C(sigma, d_mat, rho)
Emat <- diag(rep(sigma_e^2, n))
# simulate realizations
L_c <- t(chol(Cmat + Emat))
z <- rnorm(n)
y <- L_c %*% z
s <- interp(x1, x2, y, nx=300, ny=300)
image(s, col=rainbow(100))
title('OU 2d Gaussian process')
```

# Other inputs

Generally, $k(x)$ maps **distance** to **correlation**

- phylogenetic distance $\rightarrow$ phylogenetic correlation
- pedigree distance $\rightarrow$ additive genetic correlation
- distance in time $\rightarrow$ temporal correlation


# Student projects

Before fitting your model to your data: 

1. Write out model in mathematical notation (ideally \LaTeX)
2. Prior predictive simulations (do your priors make sense?)
3. Model verification (given known parameters from PPS, do you recover parameters?)

