---
title: 'Week 1 assignment: linear models'
author: "Your name here"
date: "January 6, 2015"
output: pdf_document
---

# Loading the relevant R packages

To complete these problems, you'll need some R packages loaded and ready to go.
We suggest starting with a package for plotting (ggplot2), and potentially some packages for manipulating data frames (dplyr and tidyr), depending on how you prefer to use R.

```{r}
#your code here
```

# Problem 1

You've discovered a single adult of a new species of amniote.
Though you don't know the sex, you're worried that it might lay eggs soon and you want to build something to protect them.
To do so, you need to predict the mass of this new creature's eggs.
You first weigh your amniote and find that it weighs 3000 grams.
Luckily, some researchers have just recently published a full database of aminote life history traits, which can give you some information about how amniote adult mass relates to amniote egg mass. ([Myhrvold et al. (2015)](http://www.esajournals.org/doi/abs/10.1890/15-0846R.1))

### Loading the data

Your first task will be to acquire the published dataset, which is available [here](http://www.esapubs.org/archive/ecol/E096/269/Data_Files/Amniote_Database_Aug_2015.csv).
Download the data file `Amniote_Database_Aug_2015.csv` and save it in a location that makes sense (e.g., `~/hmods/class1/data/`).
Then, you'll need to load the data with the `read.csv()` function.
Do not use `file.choose()`.

```{r}
# your code here
```

### Preparing the data

In this dataset, missing values are coded as `-999`.
We want to replace these values with `NA` which indicates missing values to R.

```{r}
# your code here
```

### Visualizing the data

We want to predict egg mass from adult body mass.
Visualize the relationship between these two variables below.
Transformation will be helpful for both variables.

```{r}
# your code here

```

### Modeling the data

Use the `lm()` function to construct a linear model that could be used to predict egg mass based on adult body mass.
(Hint: what other kinds of data transformations might be helpful prior to fitting the model?)

```{r}
# your code here

```

Evaluate the homoscedasticity and normality assumptions graphically (e.g., `plot(mymodel)`.

```{r}
# your code here

```

Are the assumptions met?
Why or why not?

*Your text answer here*

### Understanding the model

Produce a scatterplot as before, but this time add a trendline that represents the expected value of the response as a function of the predictor.

```{r}
# your code here
```

Make an image plot of the design matrix for your model (e.g., `image(t(model.matrix(m)))`):

```{r}
# your code here
```

Why does this image plot look the way it does, and what is the result of multiplying the design matrix by the vector of estimated coefficients (e.g., `model.matrix(m) %*% coef(m)`)?

*Your text answer here*

### Predicting egg mass for the new critter

Predict the egg mass for the new species in units of grams.

```{r}
# your code here
```
Besides reporting your best estimate,and provide upper and lower bounds on this estimate, in units of grams.
Remember that interval should incorporate both *predictive* uncertainty (error term in the model) and inferential uncertainty (uncertainty about the coefficients and amount of residiual error)
(Hint: there's a built-in R function that should help generate prediction intervals)

```{r}
# your code here
```

# Problem 2

A week later, you are told that the critter has been identified to be in the class Reptilia.
Use this new information to update your prediction in the code block below, commenting your code to document your thought process.

```{r}
# your code here
```

How does your new prediction compare to your prediction from Problem 1 in terms of accuracy and precision?
Is it lower or higher, and why?

*Your text answer here*

# Problem 3

Myrdahl et al. just retracted all of the adult mass data from their data set, and have advised researchers to stop using the existing adult mass data until further notice!
Given this new development, and ignoring your previous (now spurious) analysis, what's your best prediction for the critter's egg mass?
Update your prediction in the block below, commenting the code as necessary.

```{r}
# your code here
```
# Bonus Problem (optional)

When predicting the egg mass value for your unknown amniote, you probably used a built-in function in R (i.e. *predict*) to automatically generate prediction intervals.
Can you generate prediction intervals from the model without resorting to a built-in *predict* fuction?
(Hint 1: how many parameters were estimated?)
(Hint 2: can prediction intervals be simulated?)
(Hint 3: check out chapter 7 of Gelman and Hill if you have it)

```{r}
# your code here
```
