---
title: 'Week 2 assignment: likelihood'
author: "Your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

# Problem 1

Earthquakes are most forceful at their epicenters, and this force attenuates with distance. 
R has an earthquake dataset `attenu` with measurements of peak horizontal ground acceleration by distance measured at multiple recording stations, with each earthquake coded in terms of magnitude. 
See the help file for more information on this dataset by typing `?attenu`.

Your main task this week is to build one model that predicts both ground acceleration at the epicenter (distance = 0), and the acceleration by distance curve as a function of magnitude and distance from the epicenter. 
You will obtain maximum likelihood estimates for the parameters of the model using the `optim` function. 
The structure of your model is up to you. 
You can use a combination of intuition, imagination, first principles, research, and collaboration to construct your model. 
(Note: there are many possible models that one could construct!)

You will benefit from visualizing the data, specifically the relationships between the quantities of interest i.e., magnitude (`mag`), distance from epicenter (`dist`), and peak acceleration (`accel`).
Include your visualization code and plots below.

```{r}
# your code here
```

What is the equation for your model? 
Write it in \LaTeX, bounded between dollar signs (e.g., $e = mc^2$), not in normal text. 

$Your equation here$

Define all of the parameters, and explain why you formulated it in this way. 
What assumptions went into your model's construction? 

*Your text here*

Write a function called `nll` that returns the negative log likelihood for your model. 
The arguments to this function should be `theta` (the parameters), and `data` (the data).

```{r}
# Your code here
```

Use `optim` to obtain maximum likelihood estimates for your model parameters.

```{r}
# Your code here
```

Did `optim()` converge to a minimum? How do you know? 

*Your text here.*

Create a scatterplot with fitted values vs. residuals.

```{r}
# Your code here
```

What stands out in the plot of fitted values vs. residuals? 
Are you worried about any violations of assumptions? 
Why or why not? 

*Your text here.*

Plot the distance by acceleration data along with your predicted curves (starting at a distance of 0) for earthquakes of magnitude 5 and 8. 

```{r}
# Your code here
```

How do your predictions compare to the data? 
Which characteristics of the data are captured well, and which are captured poorly by your model?

*Your text here.*

# Problem 2

Pat loves to play basketball. 
You observe Pat practicing free throws at the recreation center one day. 
Pat misses 3 shots in a row. 
Generate a likelihood profile for $p$, the probability that Pat makes a free throw.

```{r}
# Your code here
```

What is your MLE for $p$, and does it make sense? Why or why not?

*Your text here*
