---
title: 'Week 7 assignment: Bayesian hierarchical models'
author: "Your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document
---

The goal of today's assignment is for you to execute a Bayesian analysis of a hierarchical model from start to finish. 
We will work with Gelman's radon data, which should be familiar to you if you have been keeping up with Gelman and Hill. 
If not, see chapter 13 for some useful content.
Radon is a radioactive gas that causes lung cancer.

The data are in the file `radon.csv`, which contains the log radon mesurements in basements and first floors of various houses in different counties across the U.S, along with a measurement of the average log uranium level in each county.
Take a moment to explore the data with the following question in mind: how do radon levels compare among floors and basements across counties in the U.S.?

```{r}
# your code here
```

Your job is to model the log radon levels as a function of floor and county using a hierarchical model. 
This is a "choose your own adventure" exercise, as you may choose among four possible models which vary in complexity. 
All models should include a basement vs. first floor effect, as this is known to affect radon levels, and beyond that the options from easiest to hardest are:

1. varying intercepts at the county level (a "random intercept" model)
2. varying intercepts along with effects of county-level mean uranium
3. varying intercepts and basement effects at the county level (a "random intercept & slope" model)
4. varying intercepts and slopes with county-level mean uranium effects

Detailed model descriptions can be found for all of these models. 
Models 1 and 2 are described in Chapter 12, and models 3 and 4 are in Chapter 13 of Gelman and Hill. 

Which model do you want to implement? 

*Your text here (1, 2, 3, or 4)*

# Model specification

Specify your model in mathematical notation below. 
Be sure to specify priors for all parameters (it's a proper Bayesian model after all, none of that `lme4` nonsense).

*Your* \LaTeX

# Prior predictive simulations

To evaluate the implications of your priors, simulate from the prior predictive distribution below.

```{r}
# your code here
```

Visualize the data in a way that makes sense for the question, and evaluate based on your knowledge (or lack thereof) of radon measurements in houses whether the priors make sense. 
It's perfectly acceptable to tune your priors until they make sense (but don't use the `log_radon` data to tune the priors!).
Unless you already know a lot about radon, do a quick internet search to get a prior sense of the mean and range of log radon levels that tend to get recorded in houses, e.g., [here](http://www.who.int/mediacentre/factsheets/fs291/en/).
The units in the original data were becquerels per cubic meter $\frac{Bq}{m^3}$ (you can't make this stuff up) and these values were log transformed to get the `log_radon` column. 

```{r}
# your code here
```


# Parameter interpretation

Write an english language description of your model parameters (at least all of the hyperparameters):

- $\sigma_y$: the residual standard deviation - accounts for unexplained variance in the log radon measurements at the observation level

- *Your other parameter descriptions here and below*


# Parameter estimation

Write a Stan file that expresses the model and priors outlined above: 

```
Your Stan code
```

Estimate the parameters, evaluate convergence, and print a summary of the posterior distribution for each parameter. 

```{r, message = FALSE}
# your code here
```


# Precision and sample size

Plot the width of the credible intervals for the county level parameters vs. sample size for each county to evaluate how precision of the estimate relates to sample size. 

```{r}
# your code here
```


# Making predictions

Unlike non-hierarchical approaches, hierarchical modeling allows you to make predictions for observed and unobserved groups. 

### Observed groups 

Suppose that we are going to sample the basement of a new house in county 70. 
Simulate from the posterior predictive distribution for the radon measurements.
Your answer should be in the form of a histogram of simulated observations.

```{r}
# your code here
```

### New groups

Suppose that we are going to sample the basement of a new house in a new county with mean log uranium level `log_u` = 0.2. 
This county is new, meaning that is not represented in the data. 
What is the posterior predictive distribution for the radon measurements? 
Your answer should be in the form of a histogram of simulated observations.

```{r}
# your code here
```

Which predictions were more precise: the predictions for the observed county or the unobserved county, and why?

```{r}
# your code here
```

*Your text here*
