---
title: 'Week 3 assignment: Bayesian inference'
author: "Your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

# Problem 1

Maure et al. (2015) conducted experiments to see how ladybird beetle diet affected interactions with parasitoid wasps.
Individual beetles were randomly assigned to one of five diet treatments:

1. 20 aphids per day
2. 2 aphids per day
3. 20 aphids per day + pollen
4. 2 aphids per day + pollen
5. pollen only

Each beetle was exposed to a female parisitoid wasp for 3 hours, which deposited eggs in the beetle, and the beetles were then fed each day and the outcome of the interaction monitored.
The authors wanted to know whether diet affected the ability to recover (i.e. not die) after the parasitoid attack.
Your task is to build a Bayesian model with Stan for beetle survival as a function of experimental treatment, estimating the probability of survival for each treatment.
To keep things simple, consider the treatments to be categorical and unordered.
The data are archived in Dryad [here](http://www.datadryad.org/resource/doi:10.5061/dryad.7fq4j).

**Important**: do not manipulate the original raw data file!
This is is error prone, leaves no written record, encourages spreadsheet farming, and is not reproducible.
And, in this case, the data are already well organized.
Read in the .xls file using the `readxl` package, then use R to massage the data as necessary.

```{r}
## Loading the raw data
library(readxl) # use the read_excel() function to load the data
# your code here

## Cleaning the data
# your code here

## Visualizing the data
# your code here
```

Write out your model using \LaTeX:

*Your \LaTeX${}$ here*

Paste your Stan model statement in the code block below, and ensure that your written model matches the notation in your Stan file:

```
Your Stan file here
```

Now, use `rstan` to fit your model.
Evaluate convergence by inspecting the $\hat{R}$ statistic and the traceplots.

```{r}
# your code here
```

Calculate posterior credible intervals, medians, means, and modes for the survival probabilities for each treatment.
Hint: posterior draws can be extracted with the `rstan::extract` function, which returns a list of arrays.

```{r}
# your code here
```

Generate a plot that shows all of the raw data along with the posterior probability distributions of recovery for each treatment:

```{r}
# your code here
```

The authors reported statistically significant differences in ladybird beetle recovery between the diet treatments.
What is your conclusion for the effect of diet on ladybird beetle recovery?

*Your text here*

# Problem 2

One of the biggest advantages of Bayesian approaches is the ease with which you can make inference on **derived parameters**.
For example, we might want to know which diet treatment gives the highest survival probability.
In one draw from the posterior distribution, we should have five estimated probabilities.
The highest probability can be recorded and stored in an object (say `best`).
We can do this for each posterior draw to produce a vector of the "best" treatments (from the beetle's perspective).
To find the posterior probability that each particular treatment is best, count the frequency of each treatment in the `best` vector, and divide by the total number of posterior draws.
Do this below using the results from problem 1, and report the posterior probabilities.

```{r}
# your code here
```

Which treatment was best? What is the probability of that treatment being the best, conditional on the data?

*Your text here*

# Problem 3

Simulate data from a normal distribution for three groups, each with a different mean.
You can assume that the standard deviations for each group are equal.
In generating data, use a design matrix to acquire the expected value for your data (somewhere in your code there should be `X %*% beta`).

```{r}
# your code here
```

Write a Stan model statement for a linear model that you can use to estimate the parameters.

```
Your Stan model here
```

Use Stan to estimate the parameters of your model.

```{r}
# your code here
```

Assess convergence of the MCMC algorithm graphically and with the Rhat statistic.

```{r}
# your code here
```

Plot the marginal posterior distributions for each parameter with a vertical line at the true value.

```{r}
# your code here
```
